---
title: "Melbourne Housing Market — Data Cleaning & Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
# This chunk sets global options for the whole document.
# include=FALSE means the chunk runs but is NOT shown in the output.
# echo=TRUE    -> show the code in the report
# warning/message=FALSE -> hide package startup noise for a cleaner report
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

# Introduction

This report documents the end-to-end analysis of the Melbourne residential
real estate market (January 2016 – March 2018). The goal is to help ordinary
residents and businesses understand how the market changes over time and across
suburbs, property types, and price ranges.

The workflow has three parts:

1. **Load** the raw data and inspect it.
2. **Clean** and pre-process the data.
3. **Explore** the data to find patterns (foundation for the Power BI dashboard).

The dataset comes from
[Kaggle](https://www.kaggle.com/datasets/anthonypino/melbourne-housing-market),
scraped from weekly results published on Domain.com.au.

# Setup

We use the **tidyverse** for data manipulation and plotting, **janitor** for
tidy column names, and **skimr** for rich summaries.

```{r packages}
library(tidyverse)
library(janitor)
library(skimr)
```

# Load the Raw Data

```{r load-data}
# read_csv() reads the raw file into a tibble (a tidy data frame).
housing_raw <- read_csv("C:/Users/farza/Job/Github/Melbourne-Housing-Market/data/raw/Melbourne_housing_FULL.csv")
```

> **Note on the path:** this `.Rmd` lives in the `R/` folder, so we use
> `../data/raw/...` — the `../` means "go up one folder" to reach the project
> root, then down into `data/raw`.

A first look at the structure:

```{r first-look}
glimpse(housing_raw)
```

The data has `r nrow(housing_raw)` rows and `r ncol(housing_raw)` columns.

# Data Understanding

Before changing anything, we inspect the data to find issues to fix.

```{r understand-missing}
# Count missing values per column, as a percentage, sorted high to low.
housing_raw |>
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "column", values_to = "missing_pct") |>
  arrange(desc(missing_pct))
```

Key issues found:

- **`Price`** (our target) is ~22% missing.
- **`BuildingArea`** (~61%) and **`YearBuilt`** (~55%) are more than half empty.
- **`YearBuilt`** contains impossible values (e.g. 1196 and 2106).
- **`Landsize`** has many zeros and extreme outliers.
- Two columns are misspelled at source: `Lattitude`, `Longtitude`.

# Data Cleaning

The cleaning pipeline runs in order. **Order matters** — for example, the date
must be parsed before we can extract the year and month from it.

## Step 1 — Standardise column names

```{r clean-names}
# clean_names() -> lower_snake_case; then we fix source misspellings by hand.
df <- housing_raw |>
  clean_names() |>
  rename(
    latitude       = lattitude,
    longitude      = longtitude,
    region_name    = regionname,
    property_count = propertycount
  )

names(df)
```

## Step 2 — Parse the date column

The raw date is text in day/month/year format (e.g. `28/01/2016`). We convert
it to a real date with `dmy()` and **verify** the conversion by comparing the
count of missing dates before and after (equal counts = every value parsed).

```{r parse-date}
na_before <- sum(is.na(df$date))
df <- df |> mutate(date = dmy(date))
na_after <- sum(is.na(df$date))

c(na_before = na_before, na_after = na_after)   # should be equal
class(df$date)                                  # "Date"
```

## Step 3 — Drop rows with a missing price

`price` is the target, so rows without it are removed (not imputed).

```{r drop-price}
rows_before <- nrow(df)
df <- df |> filter(!is.na(price))
rows_after <- nrow(df)

c(before = rows_before, after = rows_after, removed = rows_before - rows_after)
```

## Step 4 — Fix invalid `year_built` values

Melbourne was founded ~1835, so any year before 1840 (or after 2019) is a
data-entry error. We set these to `NA` rather than guessing.

```{r fix-year}
df <- df |>
  mutate(year_built = if_else(
    year_built < 1840 | year_built > 2019,
    NA_real_,
    year_built
  ))

range(df$year_built, na.rm = TRUE)   # sensible range now
```

## Step 5 — Handle zeros in `landsize` and `building_area`

A building cannot have zero area (→ `NA`). A zero land size means "not
recorded" for a house/townhouse (→ `NA`), but is expected for a unit (→ keep).

```{r fix-zeros}
df <- df |>
  mutate(
    building_area = if_else(building_area == 0, NA_real_, building_area),
    landsize      = if_else(
      landsize == 0 & type %in% c("h", "t"),
      NA_real_,
      landsize
    )
  )

c(zero_building = sum(df$building_area == 0, na.rm = TRUE),
  zero_landsize = sum(df$landsize == 0, na.rm = TRUE))
```

## Step 6 — Create helper columns for time analysis

```{r helper-cols}
df <- df |>
  mutate(
    sale_year  = year(date),
    sale_month = month(date)
  )

df |> select(date, sale_year, sale_month) |> head()
```

## Save the cleaned data

```{r save-clean}
write_csv(df, "C:/Users/farza/Job/Github/Melbourne-Housing-Market/data/processed/melbourne_housing_clean.csv")
```

The cleaned dataset has `r nrow(df)` rows ready for analysis.

# Exploratory Data Analysis

> This section will be filled in during Phase 3. Below are placeholders for the
> questions we plan to explore.

## Price distribution

```{r eda-price-dist, eval=FALSE}
# eval=FALSE means this chunk is shown but NOT run yet (a placeholder).
options(scipen = 999)
ggplot(df, aes(x = price)) +
  geom_histogram(bins = 50) +
  labs(title = "Distribution of sale prices", x = "Price (AUD)", y = "Count")
```

## Price trend over time

```{r eda-trend, eval=FALSE}
# Average price per month — to be developed in Phase 3.
```

## Most expensive suburbs

```{r eda-suburbs, eval=FALSE}
# Top suburbs by median price — to be developed in Phase 3.
```

# Summary

- The raw data was cleaned: names standardised, dates parsed, missing prices
  removed, invalid years and zero areas handled, helper columns added.
- The cleaned dataset is saved to `data/processed/` and is ready for the
  exploratory analysis and the Power BI dashboard.
